---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  eval = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.retina = 3
)
```

# batchtma R package: Methods to address batch effects

<!-- badges: start -->
<!-- badges: end -->

The goal of the batchtma package is to provide functions for batch effect-adjusting biomarker data. It implements different methods that address batch effects while retaining differences between batches that may be due to "true" underlying differences in factors that drive biomarker values.


## Installation

Currently, the development version of batchtma can be installed from
[GitHub](https://github.com/) using:

```{r}
# install.packages("remotes")  # The "remotes" package needs to be installed
library(remotes)
remotes::install_github("stopsack/batchtma")
```

Once released on [CRAN](https://CRAN.R-project.org), installation will
be possible with:

```{r}
install.packages("batchtma")
```


## Usage

Load the package:
```{r eval=TRUE}
library(batchtma)
```

Define example data with batch effects:
```{r eval=TRUE}
df <- data.frame(tma = rep(1:2, times = 10),
                 biomarker = rep(1:2, times = 10) + runif(max = 5, n = 20))
```

Run the `adjust_batch()` function to adjust for batch effects:
```{r example, eval=TRUE, fig.show="hold", out.width='40%'}
df_adjust <- adjust_batch(data = df, markers = biomarker, batch = tma, method = simple)

plot_batch(data = df, marker = biomarker, batch = tma, title = "Raw data")
plot_batch(data = df_adjust, marker = biomarker_adj2, batch = tma, title = "Adjusted data")
```

* The ["Get Started"](articles/batchtma.html) vignette has details on how to use the package. 


## Methodology

The package implements five different approaches to obtaining batch effect-adjusted biomarker values. The methods differ depending on what distributional property of batch effects they address and how they handle "true" between-batch differences. Such differences can result from confounding when batches include samples with different characteristics that are expected to lead to differences in biomarker levels. They should ideally be retained when performing batch effect adjustments.

\# | `method =`    | Approach                      | Addressed              | Retains "true" between-batch differences
---+---------------+-------------------------------+------------------------+----------------
1  | ---           | Unadjusted                    | ---                    | Yes
2  | `simple`      | Simple means                  | Means                  | No
3  | `standardize` | Standardized batch means      | Means                  | Yes
4  | `ipw`         | Inverse-probability weighting | Means                  | Yes
5  | `quantreg`    | Quantile regression           | Low and high quantiles | Yes
6  | `quantnorm`   | Quantile normalization        | All ranks              | No

* ["Get Started"](articles/batchtma.html) shows general examples on the different methods in absence and presence of confounding.


## References

* Batch means after model-based standardization for confounders (`method = "standardize"`) are partially based on a method briefly described in: *Rosner B, Cook N, Portman R, Daniels S, Falkner B. Determination of blood pressure percentiles in normal-weight children: some methodological issues. [Am J Epidemiol 2008;167(6):653-66](https://pubmed.ncbi.nlm.nih.gov/18230679).*
* Quantile normalization (`method = "quantnorm"`) is described in: *Bolstad BM, Irizarry RA, Åstrand M, Speed TP. A comparison of normalization methods for high density oligonucleotide array data based on variance and bias. [Bioinformatics 2003;19:185–193](https://pubmed.ncbi.nlm.nih.gov/12538238).*
* Further details, all methods, and suggestions for implementation and interpretation are in: *Stopsack KH et al., Batch effects in tumor biomarker studies using tissue microarrays: Extent, impact, and remediation. In preparation.*
