---
title: "Get Started: Methods to address batch effects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started: Methods to address batch effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installation

Currently, the development version of batchtma can be installed from
[GitHub](https://github.com/) using:

```{r, eval = FALSE}
# install.packages("remotes")  # The "remotes" package needs to be installed
library(remotes)
remotes::install_github("stopsack/batchtma")
```

To have vignettes like this current site be available offline via, *e.g.*, `vignette("batchtma")`, modify the last commend:

```{r, eval = FALSE}
remotes::install_github("stopsack/batchtma", build_vignettes = TRUE)
```


# Set up and plot example data

We load the batchtma package and, for convenience with data handling and plotting, the tidyverse package:

```{r, message = FALSE}
library(batchtma)
library(tidyverse)
```

We construct a toy dataset of 10 individuals (*e.g.,* tumors), each with 40 measurements (*e.g.,* cores on tissue microarrays) per batch. Unlike perhaps in the real world, for each individual tumor, we also have measurements on all the other batches.

```{r}
set.seed(123)  # for reproducibility
df <- tibble(
  # Batches:
  batch = rep(paste0("batch", LETTERS[1:4]), times = 100),
  batchnum = rep(c(1, 5, 2, 3), times = 100),
  # Participants:
  person = rep(letters[1:10], each = 40),
  # Instead of a confounder, we will use a random variable for now:
  random = runif(n = 400, min = -2, max = 2),
  # The true (usually unobservable biomarker value):
  true = rep(c(2, 2.5, 3, 5, 6, 8, 10, 12, 15, 12), each = 40),
  # The observed biomarker value with random error ("noise"):
  noisy = true + runif(max = true / 3, n = 400) * 4)

df
```

We plot the biomarker values (*y*-axis) by batch (*x*-axis), using the `plot_batch()` function. Color/shape symbolizes which participant/tumor the measurements came from. Boxes span from the 25th to the 75th percentile (interquartile range); thick lines indicate medians; asterisks indicate means.

```{r}
df %>% plot_batch(marker = noisy, batch = batch, color = person)
```


## Add batch effects

We add systematic differences between batches such that there is differential measurement error between batches in terms of mean and variance. As shown above, true values were the same beyond random error. 

```{r}
df <- df %>%
  # Multiply by batch number to differentially change variance by batch,
  # divide by mean batch number to keep overall variance the same:
  mutate(noisy_batch = noisy * batchnum / mean(batchnum) + 
           # Similarly, change mean value per batch, keeping overall mean the same:
           batchnum * 3 - mean(batchnum) * 3)

df %>% plot_batch(marker = noisy_batch, batch = batch, color = person)
```


# Perform batch effect adjustment without confounders

## Simple means

`method = simple` calculates the mean for each batch and subtracts the difference between this mean and the grand mean, such that all batches end up having a mean equivalent to the grand mean. Differences in variance between batches will remain, if they exist (as in this example).

```{r}
df %>% 
  adjust_batch(markers = noisy_batch, batch = batch, 
               method = simple) %>%
  plot_batch(marker = noisy_batch_adj2, batch = batch, color = person)
```


## Means from marginal standardization

`method = standardize` performs marginal standardization by fitting a linear regression model for biomarker values with `batch` and `confounders` as predictors, and obtains the marginal means per batch if they had the same distribution of `confounders`. Differences between these marginal means and the grand mean are subtracted as in `method = simple`. In this example, the confounder is a random variable, and the results are essentially the same as for `method = simple`.

```{r}
df %>% 
  adjust_batch(markers = noisy_batch, batch = batch, 
               method = standardize, confounders = random) %>%
  plot_batch(marker = noisy_batch_adj3, batch = batch, color = person)
```


## Inverse-probability weighting

`method = ipw` predicts the probability of a measurement being from a specific batch, given the `confounders`. Mean differences between batches are obtained from a marginal structural model with stabilized inverse-probability weights and then used as in the two preceding methods. Again, the confounder is merely a random variable in this example.

```{r}
df %>% 
  adjust_batch(markers = noisy_batch, batch = batch, 
               method = ipw, confounders = random) %>%
  plot_batch(marker = noisy_batch_adj4, batch = batch, color = person)
```


## Quantile regression

`method = quantreg`, unlike the three preceding mean-only methods, addresses two distinct properties of batches: the "offset" values (a lower quantile), potentially reflective of background signal, and an inter-quantile range, potentially reflective of the "dynamic range" of the measurement. By default, the first and third qua**r**tile are used.

```{r}
df %>% 
  adjust_batch(markers = noisy_batch, batch = batch, 
               method = quantreg, confounders = random) %>%
  plot_batch(marker = noisy_batch_adj5, batch = batch, color = person)
```

## Quantile normalization

`method = quantnorm` performs quantile normalization: values are ranked within each batch, and then each rank is assigned the mean per rank across batches. Quantile normalization ensures that all batches have near-identical biomarker distributions. However, quantile normalization does not allow for accounting for `confounders`.

```{r}
df %>% 
  adjust_batch(markers = noisy_batch, batch = batch, 
               method = quantnorm) %>%
  plot_batch(marker = noisy_batch_adj6, batch = batch, color = person)
```


# Perform batch effect adjustment in the presence of confounding

## Set up example data with confounding present

In general, confounding means that exposure and outcome share common causes. Necessary---but not sufficient---properties of a confounder are that it needs to be associated with the exposure (`batch`) and associated with the outcome(s) (`markers`). For example, if systematically higher biomarker values due to batch effects happen to occur in a tissue microarray with a higher proportion of aggressive tumors, then batch effect-adjusted biomarker values should account for the differences in proportions of aggressive tumors.

In the example, a new variable for the biomarker is generated that is affected by both confounding and batch effects. First, biomarker values with random error, `noisy`, are forced to be associated with the `confounder`, which in turn is associated with `batch`. These values, `noisy_conf`, should be considered the new "ground truth."

All following plots show color/symbol shape by `confounder`.


```{r}
set.seed(123)  # for reproducibility
df <- df %>%
  # Make confounder associated with batch:
  mutate(confounder = round(batchnum + runif(n = 200, max = 2)),
         # Make biomarker values associated with confounder:
         noisy_conf = noisy + confounder * 3 - mean(confounder) * 3)

df %>% plot_batch(marker = noisy_conf, batch = batch, color = confounder)
```


Second, batch effects for mean and variance are added.

```{r}
df <- df %>% 
  # Add batch effects to confounded biomarker values:
  mutate(noisy_conf_batch = noisy_conf * batchnum / mean(batchnum) + 
           batchnum * 3 - mean(batchnum) * 3)

df %>% plot_batch(marker = noisy_conf_batch, batch = batch, color = confounder)
```

The following examples show to what extent batch effects can be removed from `noisy_conf_batch` to recover the "ground truth," `noisy_conf`.


## Means from marginal standardization

`method = standardize` reduces batch effects while allowing batch means to differ because of the confounder. Batches with higher "true" biomarker values associated with the confounder, like batch B, retain higher means after batch effect adjustment.

```{r}
df %>% 
  adjust_batch(markers = noisy_conf_batch, batch = batch, 
               method = standardize, confounders = confounder) %>%
  plot_batch(marker = noisy_conf_batch_adj3, batch = batch, color = confounder)
```


## Inverse-probability weighting

`method = ipw` also reduces batch effects while allowing batch means to differ because of the confounder. As with marginal standardization, differences in variance between batches are not addressed.

```{r}
df %>% 
  adjust_batch(markers = noisy_conf_batch, batch = batch, 
               method = ipw, confounders = confounder) %>%
  plot_batch(marker = noisy_conf_batch_adj4, batch = batch, color = confounder)
```


## Quantile regression

`method = quantreg` reduces for batch effects of "offset" and "dynamic range" separately, accounting for differences in confounders. In the example, batch B with its high levels of the confounder retains somewhat higher values and a higher variance, unlike with the two preceding mean-only methods.

```{r}
df %>% 
  adjust_batch(markers = noisy_conf_batch, batch = batch, 
               method = quantreg, confounders = confounder) %>%
  plot_batch(marker = noisy_conf_batch_adj5, batch = batch, color = confounder)
```


## For comparison: Batch effect adjustment ignoring confounding

### Simple means

`method = simple` leads to all batches having the same mean after adjustment. In the example, this method ignores that batch B should have higher values because of higher values of the confounder.

```{r}
df %>% 
  adjust_batch(markers = noisy_conf_batch, batch = batch, 
               method = simple, confounders = confounder) %>%
  plot_batch(marker = noisy_conf_batch_adj2, batch = batch, color = confounder)
```


### Quantile regression

`method = quantreg` also ignores the confounder.

```{r}
df %>% 
  adjust_batch(markers = noisy_conf_batch, batch = batch, 
               method = quantnorm, confounders = confounder) %>%
  plot_batch(marker = noisy_conf_batch_adj6, batch = batch, color = confounder)
```


# Additional features

## Missing values

batchtma can handle missing biomarker values, which is a common phenomenon in tissue microarray studies. This situation is different from gene expression studies, where measured genes are typically measured on all batches (*e.g.*, microarrays). No coding changes are needed. If multiple biomarkers are batch-adjusted with one call of `adjust_batch(markers = c(biomarker_a, biomarker_b, biomarker_c), ...)` and different rows (e.g., tumors) have missing values for different columns (biomarkers), no values will be excluded because of missingness in another marker.


## Model diagnostics

When running `adjust_batch()`, the dataset is returned with the batch effect-adjusted variables added at the end. Yet what did `adjust_batch()` do and how well did the methods used for batch effect correction fit to the data that they were applied to? Adjustment for batch effects relies on regression models for methods `standardize`, `ipw`, and `quantreg`; in addition, `method = simple` estimates means per batch. 

The `diagnose_models()` function provides an overview of what `adjust_batch()` did, shows overview of the adjustment models, and provides access to detailed model diagnostics.

To obtain an overview of model diagnostics, `diagnose_models()` is called on the return values of `adjust_batch()`:

```{r}
# df2 is the new dataset that also contains "noisy_conf_batch_adj2":
df2 <- df %>% adjust_batch(markers = noisy_conf_batch, batch = batch, 
                           method = standardize, confounders = confounder)

# Show overview of model diagnostics:
diagnose_models(df2)
```

In this example, `method = standardize` fits a linear regression model with batch and confounders as exposure and the unadjusted biomarker as the outcome. This model can be extracted directly from `diagnose_models()`:
```{r}
fit <- diagnose_models(data = df2)$model_fits[[1]][[1]]
summary(fit)
```

The marginally standardized means per batch are the adjustment parameters estimates by this procedure. They are returned by `diagnose_models()` as `adjust_parameters`:

```{r}
diagnose_models(df2)$adjust_parameters
```

Diagnostics for regresion models can be performed on fitted models just as for any other regression model in R. For example, residuals *vs.* fitted can be plotted:

```{r}
tibble(fitted    = fitted.values(fit),
       residuals = residuals(fit)) %>%
  ggplot(mapping = aes(x = fitted, y = residuals)) +
  geom_point() +
  theme_minimal()
```
